Type Visualization for R

### 2020-06-21 paper/main.tex
paper: My edits upto 4

The tool supports three broad tasks:
\begin{itemize}
  \setlength\itemsep{0em}
  \item filtering the data set down to an amenable subset;
  \item understanding the types of function arguments and returns;
  \item comparing type signatures across different functions.
\end{itemize}
%For filtering, \typevis uses the treemap idiom
%to convey the relative scale of different packages.
For exploring and filtering the data set,
\typevis relies on treemaps to represent R packages and functions,
with node size conveying relative popularity.
Additionaly, one can use a search box to directly filter
by function name.
Once a subset of one or more functions is selected,
their type signatures are encoded as flows over types,
where flow width is proportional to frequency.
These flows can be readily compared within the same function,
or between different functions.

\typevis can assist language designers during multiple phases of development.
For example, exploratory data analysis can identify unexpected edge cases,
or weed out language designs that are incompatible with existing programs.
\typevis was built in support of a new gradual type system for R,
whose adoption depends on integrating well with existing R code.
However, the visualization is not limited to R and can be used to
analyze type usage in any language where similar data is available.

Supplemental material, including the data and source code of
\typevis, is freely available: \todo{Fill with archival link}

@@

R is a dynamically typed programming language widely used for
statistical computing. Unlike statically typed languages,
R programs do not contain type annotations
and never pass through a typechecker.
At run-time,
functions may be called with values of arbitrary type,
and may return values of arbitrary type.\footnote{
  We use the word ``type'' to mean what others may call a ``type tag.''
  This is a run-time concept---we do not assume a static type system
  because there is none. Indeed, the goal is to help design one!
  A ``type signature'' in this context is one particular configuration
  of argument and return types a function is called with at run-time.
}

Being a dynamically typed language most actively used in a single domain,
R is a good target for a gradual type system.
For statically typed languages, the type system is usually
an inseparable part of the language. Since they are designed
at the same time, the (static) type system precedes any code written
in the language.
A gradual type system, by contrast,
permits the integration of static and dynamic typing.
Such a type system is capable of seamlessly mixing typed and untyped regions of code.
Often, a gradual type system is designed atop
an existing dynamically typed language.
Since many R programs already exist,
a successful gradual type system must be capable of
typechecking idiomatic R code.
\typevis aims to clarify how existing R programs
use types to assist in this task.
While this is our motivation, the visualization's use
is not necessarily limited to this application.
Anyone who has an interest in how the real R programs use types
may find it useful.

We are not aware of any previous work that applies
visualization techniques for the purposes of programming language design.
However, there are existing publications on understanding
software systems through visualization.
Bohnet et al.~\cite{bohnet:2009} visualize massive program execution traces
to assist developers in code comprehension of large software systems.
Their tool applies pruning and summarization to reduce trace size.
The visualization focuses on call topology,
describing the structure of function invocations,
rather than types.
Telea et al.~\cite{telea:2009} also develop a methodology for displaying call structure.
However, they visualize call graphs of systems in their entirety,
rather than individual execution traces.
Our visualization can be thought of as a hybrid of the two,
investigating the system as a whole,
but doing so via execution traces.
In addition to standard call graphs,
LaToza and Myers~\cite{latoza:2011} adorn nodes and edges with extra information,
allowing the user gain further understanding of the system
under examination.
This includes indicating methods that come from the same class---a
basic form of type information.

@@

\todo{Alexi's paper says 412, but I see more packages in our data.}

The data is provided by Alexi Turcotte:
it consists of traces from 412 packages
containing over $760,000$ lines of R code
and $534,000$ lines of native code.
All libraries were obtained from CRAN,
the most popular R code repository.
Packages hosted on CRAN must satisfy certain criteria
to ensure they are high quality.
Additionally, the subset of 412 packages chosen here
all had code coverage above 65\%.\todo{I don't understand what subset? Should we remove this line?}

Each execution trace was recorded by a dynamic analysis
running the test, example, and vignette code of each package.
A vignette is a form of documentation that weaves
together prose and executable code.
An execution trace is a list of ``raw'' function call signatures
containing, among other attributes, the function name
and type tags for each argument and return value.
For example, for the function call \code{!(FALSE)},
the recorded type signature would be \code{logical->logical},
meaning that function \code{!} took a logical value and returned a logical value.
The type tags are assigned by the dynamic analysis and
refine over actual type tags in R; for example, the analysis
assigns scalar tag \code{double} to a singleton vector of double values
because R does not have native support for scalars.
Our data set consists of a table of reduced execution traces:
because \typevis is designed for the analysis of types, not all information
in the traces is relevant.
%that have been processed to reduce some fields.
%Traces were generated via a dynamic analysis
%that instruments the R virtual machine with hooks
%that intercept the required data at run-time.

While \typevis has only been instantiated with
data from R execution traces, the architecture is not
limited to R.
Any dynamically typed language %with the right run-time
%hooks could generate data for the tool.
(or a statically typed language with run-time tag information)
could be instrumented to generate data for the tool.
As long as execution trace information adheres to
the same database schema, \typevis will
work properly without any modifications.

@@

We interviewed the primary researcher developing R's gradual type system
to inventory relevant domain-specific tasks.
Additionally, we sought feedback throughout the creation of \typevis
to ensure it was adequately satisfying requirements.
The domain-specific goals can be reduced to:

\begin{enumerate}
\item {\bf Find Function.} Quickly search for a particular function and identify basic information. This includes how often the function is called and which package contains it.
\item {\bf Determine Types.} Given a specific function, determine the argument and return types of the function. Additionally, identify the frequency of a type signature. A user should be able to infer, for example, if the function is polymorphic or if it's a predicate.
\item {\bf Compare Signatures.} Compare the occurrence of type signatures within a specific function, but also across many different ones. Users must be able to isolate individual signatures and see them in the context of other signatures and functions.
\end{enumerate}

These goals can be classified using an abstract task analysis framework.
See table~\ref{tab:tasks} for a categorization of these domain tasks
the mid-level and low-level abstract task taxonomy
established by Brehmer and Munzner~\cite{brehmer:2013}.

>>>>

The tool supports three broad tasks:
\begin{itemize}
  \setlength\itemsep{0em}
  \item filtering the data set down to an amenable subset;
  \item understanding the types of function arguments and returns;
  \item comparing type signatures across different functions.
\end{itemize}
%For filtering, \typevis uses the treemap idiom
%to convey the relative scale of different packages.
For exploring and filtering the data set,
\typevis relies on treemaps to represent R packages and functions,
with node size conveying relative popularity.
Additionally, one can search directly for a function by name.
Once a subset of one or more functions is selected,
their type signatures are encoded as flows over types,
where flow width is proportional to frequency.
These flows can be readily compared within the same function,
or between different functions.

\typevis can assist language designers during multiple phases of development.
For example, exploratory data analysis can identify unexpected edge cases,
or weed out language designs that are incompatible with existing programs.
\typevis was built in support of a new gradual type system for R,
whose adoption depends on integrating well with existing code.
However, the visualization is not limited to R and can be used to
analyze type usage in any language where similar data is available.

Supplemental material, including the data and source code of
\typevis, is freely available: \todo{Add archival link}

@@

R is a lazy, multi-paradigm, dynamically-typed programming language
widely used for statistical computing~\cite{morandat:2012}.
Unlike statically-typed languages,
R code contains no type annotations
and never passes through a typechecker.
At run-time,
functions may be called with values of arbitrary type,
and may return values of arbitrary type.\footnote{
  We use the word ``type'' to mean what others may call a ``type tag.''
  This is a run-time concept---we do not assume a static type system
  because there is none. Indeed, the goal is to help design one!
  A ``type signature'' in this context is one particular configuration
  of argument and return types a function is called with at run-time.
}

Being a dynamically-typed language
with widespread use in a single domain,
R is a good target for a gradual type system.
A gradual type system permits the integration of
static and dynamic typing~\cite{siek:2006,tobin-hochstadt:2006}.
Such a type system is capable of seamlessly mixing typed and untyped regions of code.

For statically-typed languages, the type system is usually
an inseparable part of the language. Since they are designed
at the same time, the static type system precedes any code written
in the language.
Gradual type systems, by contrast, are often designed atop
an existing dynamically-typed language~\cite{tobin-hochstadt:2016}.
Many R programs already exist,
so a successful gradual type system must
work well with extant code.

We are not aware of any previous work that applies
visualization techniques for the purpose of programming language design.
However, there are existing publications on understanding
software systems through visualization.
Bohnet et al.~\cite{bohnet:2009} visualize massive program execution traces
to improve developer comprehension of large software systems.
Their tool applies pruning and summarization techniques to reduce trace size.
They focus on call topology,
describing the structure of function invocations,
rather than types.
Telea et al.~\cite{telea:2009} also develop a methodology for displaying call structure.
However, they visualize call graphs of systems in their entirety,
rather than individual execution traces.
\typevis has elements of both,
investigating the system as a whole,
but doing so via execution traces.
In addition to standard call graphs,
LaToza and Myers~\cite{latoza:2011} adorn nodes and edges with extra information,
allowing the user to gain further understanding of the system
under examination.
This includes grouping methods that come from the same class---a
basic form of type data.

@@

The data set we used comes from
Turcotte et al.~\cite{turcotte:2020}
and consists of traces from 412 packages,
containing over $760,000$ lines of R code
and $534,000$ lines of native code.
All libraries were obtained from CRAN,
the most popular R code repository.
Packages hosted on CRAN are curated and satisfy
certain quality standards.
Out of the over $15,000$ packages
distributed on CRAN,
the subset of 412 packages considered here
all had code coverage above 65\%.
\todo{Confirm with Alexi this paragraph. Get proper BibTeX entry.}

Execution traces were recorded by running a dynamic analysis
over the test, example, and vignette code of each package.
A vignette is a form of documentation that weaves
together prose and executable code.
An execution trace is a list of function type signatures
containing, among other attributes, the function name
and type tags for each argument and return value.
For example, the function call {\tt !(FALSE)}
would be recorded as the type signature {\tt logical $\to$ logical},
meaning that the function {\tt !} took a logical value
and returned a logical value.\footnote{
  In R, a logical value is a boolean.
}
The type tags are assigned by the dynamic analysis
and refine the actual type tags in R.
For instance, the analysis
assigns the scalar tag {\tt double} to a singleton vector of double values
because R does not have native support for scalars.
The data set is a table of reduced execution traces.
Because \typevis was designed for the analysis of types,
not all information in the traces is relevant.
%that have been processed to reduce some fields.
%Traces were generated via a dynamic analysis
%that instruments the R virtual machine with hooks
%that intercept the required data at run-time.

While \typevis has only been instantiated with
data from R execution traces, the architecture is not limited to R.
Any dynamically-typed language, %with the right run-time
%hooks could generate data for the tool.
or statically-typed language with run-time tag information,
could be instrumented to generate data for the tool.
As long as the trace output adheres to
the correct schema, \typevis will work properly without any modifications.

@@

We interviewed the primary researcher developing a gradual type system for R
to inventory relevant domain-specific tasks.
Additionally, we sought feedback throughout the creation of \typevis
to ensure it was adequately meeting requirements.
The domain-specific goals identified were:

\begin{enumerate}
\item {\bf Find Function.} Quickly search for a particular function and identify basic information. This includes how often the function is called and which package contains it.
\item {\bf Determine Types.} Given a specific function, determine the argument types, return types, and identify the frequency of specific type signatures. A user should be able to readily infer, for example, if a function is polymorphic or if it is a predicate.
\item {\bf Compare Signatures.} Compare the occurrence of type signatures within a specific function, and across many different functions. Users must be able to isolate individual signatures and see them in the context of others.
\end{enumerate}

These goals can be classified using an abstract task analysis framework.
See table~\ref{tab:tasks} for a categorization of these domain tasks
with the mid-level and low-level abstract task taxonomy
established by Brehmer and Munzner~\cite{brehmer:2013}.
